---
title: "Assignment 10: Data Scraping"
author: Hannah Wudke
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping.

## Directions

1.  Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up

1.  Set up your session:

-   Load the packages `tidyverse`, `rvest`, and any others you end up using.
-   Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(here)
library(dplyr)
library(rvest)
library(ggplot2)
library(ggridges)

here()
```

2.  We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2023 Municipal Local Water Supply Plan (LWSP):

-   Navigate to <https://www.ncwater.org/WUDC/app/LWSP/search.php>
-   Scroll down and select the LWSP link next to Durham Municipality.
-   Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2023>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
water.webpage <- 
  read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2023')
```

3.  The data we want to collect are listed below:

-   From the "1. System Information" section:

-   Water system name

-   PWSID

-   Ownership

-   From the "3. Water Supply Sources" section:

-   Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

> HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
Water.System.Name <- water.webpage %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
Water.System.Name

PWSID <- water.webpage %>% html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% html_text()
PWSID

Ownership <- water.webpage %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
Ownership

MGD <- water.webpage %>% html_nodes("th~ td+ td") %>% html_text()
MGD

```

4.  Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

> TIP: Use `rep()` to repeat a value when creating a dataframe.

> NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5.  Create a line plot of the maximum daily withdrawals across the months for 2023, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
DurhamWater <- data.frame("Month" = rep(1:12), "Year" = rep(2023,12),
                          "Max.Day.Use" = as.numeric(MGD))

DurhamWater <- DurhamWater %>% 
  mutate(Ownership = !!Ownership,
         PWSID = !!PWSID,
         Water.System.Name = !!Water.System.Name, Date = my(paste(Month, "-", Year)))
#5 
Durham.Water.Plot <- ggplot(DurhamWater,
            aes(x=Date,y=Max.Day.Use)) + geom_line() + geom_smooth(method="loess",
            se=FALSE) + labs(title = paste("Water usage data for",
            Water.System.Name), subtitle = Ownership, y="Withdrawal", x="Date") 

print(Durham.Water.Plot)

```

6.  Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data, returning a dataframe. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
scrape.fxn <- function(Year, PWSID){
  Water.Website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',
                                    PWSID, '&year=', Year))
  
Ownership.Tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
MGD.Tag <- 'th~ td+ td'
PWSID.Tag <- 'td tr:nth-child(1) td:nth-child(5)'
Water.System.Name.Tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'

The.Ownership <- Water.Website %>% html_nodes(Ownership.Tag) %>% html_text()
The.MGD <- Water.Website %>% html_nodes(MGD.Tag) %>% html_text()
The.PWSID <- Water.Website %>% html_nodes(PWSID.Tag) %>% html_text()
The.Water.System.Name <- Water.Website %>% html_nodes(Water.System.Name.Tag) %>% html_text()

The.Dataframe <- data.frame("Month" = rep(1:12), "Year" = rep(Year,12),
  "Average.Withdrawals.MGD" = as.numeric(The.MGD)) %>% mutate(Ownership = !!The.Ownership,
        Water.System.Name = !!The.Water.System.Name, PWSID = !!The.PWSID, 
        Date = my(paste(Month,"-",Year)))

Sys.sleep(3)

return(The.Dataframe)
}


```

7.  Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
Durham2015 <- scrape.fxn(2015, '03-32-010')

Durham.Water.Plot.2015 <- ggplot(Durham2015,
            aes(x=Date,
          y=Average.Withdrawals.MGD)) + geom_line() + geom_smooth(method="loess",
            se=FALSE) + labs(title = paste("Water usage data for",
            Water.System.Name), subtitle = Ownership, y="Withdrawal", x="Date") 
print(Durham.Water.Plot.2015)
```

8.  Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
Asheville2015 <- scrape.fxn(2015, '01-11-010')
view(Asheville2015)

Asheville.Durham <- bind_rows(Asheville2015, Durham2015)

 Asheville.Durham.Plot <- ggplot(Asheville.Durham,
            aes(x=Date,y=Average.Withdrawals.MGD,
            color=Water.System.Name)) + geom_line() + geom_smooth(method="loess",
            se=FALSE) + labs(title = paste("Water usage data for Asheville and Durham"), 
                          subtitle = Ownership, y="Withdrawal", x="Date")
print(Asheville.Durham.Plot)
 
```

9.  Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2018 thru 2022.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one.

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
Asheville2018 <- scrape.fxn(2018, '01-11-010')
Asheville2019 <- scrape.fxn(2019, '01-11-010')
Asheville2020 <- scrape.fxn(2020, '01-11-010')
Asheville2021 <- scrape.fxn(2021, '01-11-010')
Asheville2022 <- scrape.fxn(2022, '01-11-010')

Asheville.Years <- bind_rows(Asheville2018, Asheville2019, Asheville2020,
                             Asheville2021, Asheville2022)
 Asheville.Years.Plot <- ggplot(Asheville.Years,
  aes(x=Date,y=Average.Withdrawals.MGD)) + geom_line() + geom_smooth(method="loess",
            se=FALSE) + labs(title = paste("Water usage data for Asheville"), 
                             subtitle = Ownership, y="Withdrawal", x="Date")
print(Asheville.Years.Plot)
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time? \> Answer: Yes, water usage increases from 2018 to 2020, decreases for a year, likely due to the pandemic, and then steadily increases again towards 2022. \>
